---
title: "EXPLORATORY ANALYSES OF TOOLIBIN AND WALBYRING WATERBIRD COUNTS and INVERTEBRATES INCLUDING 2021-22"
author: "Adrian Pinder"
date: 'Date and time: `r Sys.time()`'
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, include = FALSE}
packages <- c("gridExtra","dendextend","bibliometrix","data.table","mtcars","dplyr","scatterplot3d","plotly","RColorBrewer","stringr","rstudioapi","patchwork","glue","ClassDiscovery","vegan","tidyverse","lubridate","ggplot2","ggpubr","stringi","factoextra", "zoo","cluster","rgl","reshape2","ggthemes","data.table")
lapply(packages, require, character.only=TRUE)
```

Software: Rstudio version: `r rstudioapi::versionInfo()$version`; R version: `r getRversion()`

```{r echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = TRUE)
options(scipen=999)
set.seed(12345)
```

# BACKGROUND

Analyses below are for a report by Pinder et al.on 2021/22 data.

Lake Toolibin filled in winter 2021 from rainfall in July/August. Surveys of waterbirds were undertaken across almost all of Lake Toolibin and the downstream Lake Walbyring on Nov 1 and 2 2021 (3 months after filling) and 10-11 Jan 2022 (4.5 months after filling) by boat or (for Jan 2022 at Walbyring) on foot by traversing the perimeter of the lake. Invertebrate surveys were undertaken at the same time using the methods of Cale and Pinder (2019).

Cale and Pinder (2019) analysed patterns in waterbird composition from 1) the 1980s (Jaensch et al., 1985), a survey undertaken in 1996 by Halse <i>et al.</i> (2000) and two 2017 surveys following a summer fill event commencing 9 February 2017. They also analysed invertebrate data collected from both lakes in 1992 (Doupe and Horwitz, 1995), 1996 (Halse <i>et al.</i>, 2000) and 2017. This report extends the Cale and Pinder (2019) analyses of waterbirds and invertebrates to include the 2021/22
data.

# WATERBIRD ANALYSES

The waterbird data input files include depth at time of survey, visit date, period (1980s, 1996, 2017 and 2021/22), an indication (A = ALL_SPECIES) of whether the observers thought they had adequately surveyed the whole of the lake and seen all species (A), whether they were considered pre-salinity reference communities by Froend et al (FHS_Ref), or as an expanded reference set of sites in these analysis, whether they were used as 'comprehensive surveys in these analyses (All_Ref = Y), and species richness.

```{r}
rawdata <- read.csv("./Newdata/Allsurveys18 3.csv", row.names = 1) 
#retain 'rawdata' as an unmodified waterbird import
rawdata$VISIT_DATE <- as.Date(rawdata$VISIT_DATE, "%d/%m/%Y") #convert waterbird survey d
```

## Graphs of richness by Period for Toolibin and Walbyring

```{r  fig.align = 'center'}}
rawdata3 <- rawdata
rawdata3$wetland <- str_sub(row.names(rawdata3),1,1) #created new var for Toolibin or Walbyring from survey codes
```

```{r}
#Toolibin scatterplot
rawdata3.T <- filter(rawdata3, wetland == "T") #extract just Toolibin data
rawdata3.T$colour <- rawdata3.T$All_Ref #new variable (colour) to indicate whether survey is considered reference (Y and N) or is part of the 1996+ monitoring (M) surveys
rawdata3.T$colour <- gsub('N', 'Blue', rawdata3.T$colour)
rawdata3.T$colour <- gsub('Y', 'Red', rawdata3.T$colour)
rawdata3.T$colour <- gsub('M', 'Black', rawdata3.T$colour)

ggplot(data = rawdata3.T, aes(x = Period, y = RICHNESS)) + geom_jitter(aes(colour = colour, alpha = 0.2), size=8, width = 0.05, height = 0.05) + xlab("Period") + ylab("Number of waterbird species") + theme(axis.title.x = element_text(margin=margin(t=20))) + theme(axis.title.y = element_text(margin=margin(r=20))) +theme(axis.text=element_text(size=10), axis.title=element_text(size=14,face="bold")) + theme(legend.position="none") + scale_color_manual(values = c("Blue", "Grey", "Red"))
```

```{r}
#Walbyring scatterplot
rawdata3.W <- filter(rawdata3, wetland == "W")
rawdata3.W$colour <- rawdata3.W$All_Ref #new variable (colour) for depth and completeness of surveys
ggplot(data = rawdata3.W, aes(x = Period, y = RICHNESS)) + geom_jitter(aes(colour = colour, alpha = 0.2), size=8, width = 0.05, height = 0.05) + xlab("Period") + ylab("Number of waterbird species") + theme(axis.title.x = element_text(margin=margin(t=20))) + theme(axis.title.y = element_text(margin=margin(r=20))) +theme(axis.text=element_text(size=10), axis.title=element_text(size=14,face="bold")) + theme(legend.position="none") + scale_color_manual(values = c("Blue", "Grey", "Red"))
```


## Multivariate analyses including all surveys (from both lakes) 

All multivariate analyses are performed on Helinger transformed abundance data (within species abundances scaled to values between 0 and 1 and square root transformed). Surveys with richness <=1 removed to deal with an outlier identified in the 2019 analyses. Also removed the 1975 survey as this was not quantitative.

```{r}
comdata <- subset(rawdata, RICHNESS > 1)
comdata <- comdata[!rownames(comdata) == "T1075", ]
comdata.fact <- comdata[, 1:8] #add factorsfactors
comdata.fact$Survey <- row.names(comdata.fact)
comdata <- comdata[, -(1:8)] #remove factors from data file
comdata[1:5, 1:5] #check first five lines & columns of data file
comdata[is.na(comdata)]<-0 # replace na with 0
comdata.pa<-decostand(comdata,"pa") #create presence/absence matrix
comdata.hel<-decostand(comdata,"hellinger") # create helinger transformed matrix
```

```{r include=FALSE}
mds1 <- metaMDS(comdata.hel, distance ="bray", k=2) #nMDS in 2 dimensions
```

This two-dimensional ordination had unacceptably high stress value (`r round(mds1$stress, digits=2)`) so a 3 dimensional ordination was run.

```{r include=FALSE}
mds2 <- metaMDS(comdata.hel, distance = "bray", k=3) #nMDS in 3 dimensions
```

```{r}
#Extract mds2 coordinates and create new dataframe with coordinates, survey, and survey period
mds2.pts <- as.data.frame(mds2$points) #extract 2D ordination coordinates (points)
mds2.pts$Survey <- row.names(mds2.pts)
mds2.pts$Period2 <- comdata.fact[match(mds2.pts$Survey, comdata.fact$Survey), 3] #column '2' has 'Period'
```

Below is a plot of axes 1 v 2 of the 3D ordination, showing all of the 1996 to 2022 surveys on the edge of the 1975 and 1980s surveys.

```{r fig.align = 'center'}
#first set axis limits based on maximum values of coordinates
Xmin <- min(mds2.pts$MDS1)*1.2 
Xmax <- max(mds2.pts$MDS1)*1.2
Ymin <- min(mds2.pts$MDS2)*1.2
Ymax <- max(mds2.pts$MDS2)*1.2
ggplot(mds2.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Period2), size=5) + coord_fixed(ratio = 1)
```

## Multivariate analyses with only Lake Toolibin data

When just the Toolibin Surveys are considered, an ordination plot still shows the 1996 to 2022 surveys on the edge of the 1980s point cloud.

```{r}
comdataRT <- rawdata[rawdata$RICHNESS > 1, ]
comdataRT <- comdataRT[!rownames(comdataRT) == "T1075", ]
comdataRT[is.na(comdataRT)] <- 0 # replace na with 0
comdataRT <- comdataRT[grep("^T", row.names(comdataRT)),] #extracts Toolibin
comdataRT.fact <- comdataRT[, 1:8]
comdataRT.fact$Survey <- row.names(comdataRT.fact)
comdataRT.fact$YM <- format(as.Date(comdataRT.fact$VISIT_DATE), "%m-%y")
comdataRT.fact$Depthcat <- if_else(comdataRT.fact$DEPTH >0.599, "D", "S") #D = deep, S = shallow
comdataRT.fact$AvN <- paste(comdataRT.fact$ALL_SPECIES, comdataRT.fact$Depthcat) #new variable (colour) for depth and completeness of surveys
comdataRT <- comdataRT[, -(1:8)] #remove factors from data file
comdataRT <- comdataRT[, colSums(comdataRT) != 0]
comdataRT.pa <-decostand(comdataRT,"pa") #create presence/absence matrix
comdataRT.hel <-decostand(comdataRT,"hellinger") # create helinger transformed matrix
row.names(comdataRT.hel) <- comdataRT.fact$VISIT_DATE
comdataRT.sqrt <- sqrt(comdataRT)
```

```{r include=FALSE}
mds6<-metaMDS(comdataRT.hel, distance="bray", k=3) #nMDS in 3 dimensions
```

```{r}
#Extract mds points and add factors
  mds6.pts <- as.data.frame(mds6$points) #extract 2D ordination coordinates (points)
  mds6.pts$Survey <- row.names(comdataRT.fact) #use survey period as row.names
  mds6.pts$YM <- comdataRT.fact$YM #add YearMonth for labelling
  mds6.pts$Period2 <- comdataRT.fact$Period2
  mds6.pts$All_Ref <- comdataRT.fact$All_Ref
  mds6.pts$Depth <- comdataRT.fact$DEPTH
  mds6.pts$AvN <- comdataRT.fact$AvN
  #mds6.pts$Period <- comdataRT.fact[match(row.names(mds6.pts), row.names(comdataRT.fact)), 2]
  #mds6.pts$Depth <- comdataRT.fact[match(row.names(mds6.pts), row.names(comdataRT.fact)), 4]
```

```{r fig.align = 'center'}
Xmin <- min(mds6.pts$MDS1)*1.2
Xmax <- max(mds6.pts$MDS1)*1.2
Ymin <- min(mds6.pts$MDS2)*1.2
Ymax <- max(mds6.pts$MDS2)*1.2
ggplot(mds6.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Period2), size=6, alpha=0.8) + coord_fixed(ratio = 1) + scale_color_manual(values = c("dodgerblue4", "deepskyblue2", "cadetblue1", "Grey","Red")) + labs(colour = "Period") 
```

```{r eval = FALSE}
#cluster analyses of Toolibin surveys
plot(comdataRT.upgma <- hclust(vegdist(comdataRT.hel), method="aver"), hang=1)
tabasco(comdataRT.hel,comdataRT.upgma)
```

## Multivariate analyses with only Lake Walbyring data

```{r}
comdataRW <- rawdata[rawdata$RICHNESS > 4, ] #extract surveys with richness >4
comdataRW[is.na(comdataRW)] <- 0 # replace na with 0
comdataRW <- comdataRW[grep("^W", row.names(comdataRW)),] #extracts Walbyring
comdataRW.fact <- comdataRW[, 1:8] #extracts factors as anew file
comdataRW.fact$Survey <- row.names(comdataRW.fact)
comdataRW.fact$YM <- format(as.Date(comdataRW.fact$VISIT_DATE), "%m-%y")
comdataRW <- comdataRW[, -(1:8)] #remove factors from data file
comdataRW <- comdataRW[, colSums(comdataRW) != 0]
comdataRW.pa <-decostand(comdataRW,"pa") #create presence/absence matrix
comdataRW.hel <-decostand(comdataRW,"hellinger") # create helinger transformed matrix
row.names(comdataRW.hel) <- comdataRW.fact$VISIT_DATE
```

```{r include=FALSE}
mds6<-metaMDS(comdataRW.hel, distance="bray", k=3) #nMDS in 3 dimensions
```

```{r}
#Extract mds points and add factors
  mds6.pts <- as.data.frame(mds6$points) #extract 2D ordination coordinates (points)
  mds6.pts$Survey <- row.names(comdataRW.fact) #use survey period as row.names
  mds6.pts$YM <- comdataRW.fact$YM #add YearMonth for labelling
  mds6.pts$Period2 <- comdataRW.fact$Period2
  mds6.pts$All_Ref <- comdataRW.fact$All_Ref
  mds6.pts$Depth <- comdataRW.fact$DEPTH
  mds6.pts$AvN <- comdataRW.fact$AvN
  #mds6.pts$Period <- comdataRW.fact[match(row.names(mds6.pts), row.names(comdataRW.fact)), 2]
  #mds6.pts$Depth <- comdataRW.fact[match(row.names(mds6.pts), row.names(comdataRW.fact)), 4]
```

Axes 1 v 2 of a 3D ordination with stress
`r round(mds6$stress, digits=3)` showing the 1 x 1996, 2 x 2017 and 2 x
2021-22 surveys on the edge of the 1980s surveys.

```{r fig.align = 'center'}
Xmin <- min(mds6.pts$MDS1)*1.2
Xmax <- max(mds6.pts$MDS1)*1.2
Ymin <- min(mds6.pts$MDS2)*1.2
Ymax <- max(mds6.pts$MDS2)*1.2

ggplot(mds6.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Period2), size=6, alpha=0.8) + coord_fixed(ratio = 1) + scale_color_manual(values = c("dodgerblue4", "deepskyblue2", "cadetblue1", "Grey","Red")) + labs(colour = "Period") 
```

```{r eval = FALSE}
plot(comdataRW.upgma <- hclust(vegdist(comdataRW.hel), method="aver"), hang=1)
tabasco(comdataRW.hel,comdataRW.upgma)
```

## Heatmap of species abundances by reference survey for Toolibin Lake. Note the y axis is not sorting by date - need to fix.
```{r}
comdataHM <- rawdata[rawdata$All_Ref %in% c("Y","M"), ] #selects reference surveys
comdataHM<- comdataHM[!rownames(comdataHM) == "T1075", ] #removes 1975 presence absence data 
comdataHM <- comdataHM[grep("^T", row.names(comdataHM)),] #extracts Toolibin
comdataHM[is.na(comdataHM)] <- 0 # replace na with 0
row.names(comdataHM) <- comdataHM$VISIT_DATE
comdataHM <- comdataHM[, -(1:8)]
comdataHM <- comdataHM[, colSums(comdataHM) != 0]
comdataHM.hel <-decostand(comdataHM,"hellinger") # create helinger transformed matrix
melt_comdataHM.hel <- melt(as.matrix(comdataHM.hel))
ggplot(melt_comdataHM.hel, aes(Var2, Var1)) + geom_tile(aes(fill = value), colour = "white") + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_gradient(low = "lightblue", high = "red") + labs(x="Species",y="Survey") + theme(axis.text.y = element_text(colour = c("black", "black","black","black","black","black","black","black","black","black","black","blue", "blue", "blue", "blue", "blue", "blue", "blue")))
```

## Simper analysis of just the reference surveys versus 1996+ surveys for Lake Toolibin
```{r}
comdataRTA <- rawdata[rawdata$All_Ref %in% c("Y","M"), ]
comdataRTA[is.na(comdataRTA)]<-0 # replace na with 0
comdataRTA <- comdataRTA[grep("^T", row.names(comdataRTA)),] #extracts Toolibin
comdataRTA <- comdataRTA[!rownames(comdataRTA) == "T1075", ] #removes 1975 presence absence data 
comdataRTA.fact <- comdataRTA[, 1:8]
comdataRTA <- comdataRTA[, -(1:8)] #remove factors from data file
comdataRTA <- comdataRTA[, colSums(comdataRTA) != 0]
comdataRTA.hel<-decostand(comdataRTA,"hellinger") # create helinger transformed matrix
#simper on transformed data
comdataRTA.Simper <- simper(comdataRTA.hel, comdataRTA.fact$All_Ref)
summary(comdataRTA.Simper)
#simper on untransformed data
comdataRTA.Simper <- simper(comdataRTA, comdataRTA.fact$All_Ref)
summary(comdataRTA.Simper)
```

## Plots of abundance for target species in the management plan.
```{r}
comdataRTA.2 <- as.data.table(comdataRTA)
comdataRTA.2$Period <- comdataRTA.fact$Period2
comdataRTA.2$Period <- ifelse(comdataRTA.2$Period %in% c("1996","2017","2021/22"), comdataRTA.2$Period, "Reference")
```

```{r}
#create new columns for specifying ggplot pch values based on presence/absence of each of the indicator species
comdataRTA.2[, FD.char := ifelse(Freckled.Duck == "0", 1, 16)]
comdataRTA.2$FD.char <- as.factor(comdataRTA.2$FD.char)
comdataRTA.2$Fd.char <- as.factor(comdataRTA.2$FD.char)

comdataRTA.2[, AS.char := ifelse(Australian.Shelduck == "0", 1, 16)]
comdataRTA.2$AS.char <- as.factor(comdataRTA.2$AS.char)
comdataRTA.2$AS.char <- as.factor(comdataRTA.2$AS.char)

comdataRTA.2[, PD.char := ifelse(Pink.eared.Duck == "0", 1, 16)]
comdataRTA.2$PD.char <- as.factor(comdataRTA.2$PD.char)
comdataRTA.2$PD.char <- as.factor(comdataRTA.2$PD.char)

comdataRTA.2[, EC.char := ifelse(Eurasian.Coot == "0", 1, 16)]
comdataRTA.2$EC.char <- as.factor(comdataRTA.2$EC.char)
comdataRTA.2$EC.char <- as.factor(comdataRTA.2$EC.char)

comdataRTA.2[, BS.char := ifelse(Black.winged.Stilt == "0", 1, 16)]
comdataRTA.2$BS.char <- as.factor(comdataRTA.2$BS.char)
comdataRTA.2$BS.char <- as.factor(comdataRTA.2$BS.char)

comdataRTA.2[, GT.char := ifelse(Grey.Teal == "0", 1, 16)]
comdataRTA.2$GT.char <- as.factor(comdataRTA.2$GT.char)
comdataRTA.2$GT.char <- as.factor(comdataRTA.2$GT.char)
```

```{r fig.align = 'center'}}
max_y <- max(comdataRTA.2$Freckled.Duck)
FD <- ggplot(comdataRTA.2, aes(x = Period, y = Freckled.Duck, alpha = 0.2, shape = factor(FD.char))) + 
  scale_shape_manual(values = c(1, 16)) +
  geom_jitter(width = 0.05, size = 7, colour="blue") + 
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) + 
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Freckled Duck") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

max_y <- max(comdataRTA.2$Australian.Shelduck)
AS <- ggplot(comdataRTA.2, aes(x = Period, y = Australian.Shelduck, alpha = 0.2, shape = factor(AS.char))) + 
  scale_shape_manual(values = c(1, 16)) +
  geom_jitter(width = 0.05, height = 0, size = 7, colour="blue") + 
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) + 
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Australian Shelduck") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

max_y <- max(comdataRTA.2$Pink.eared.Duck)
PD <- ggplot(comdataRTA.2, aes(x = Period, y = Pink.eared.Duck, alpha = 0.2, shape = factor(PD.char))) + 
  scale_shape_manual(values = c(1, 16)) +
  geom_jitter(width = 0.05, height = 0, size = 7, colour="blue") +  
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) + 
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Pink-eared Duck") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

max_y <- max(comdataRTA.2$Eurasian.Coot)
EC <- ggplot(comdataRTA.2, aes(x = Period, y = Eurasian.Coot, alpha = 0.2, shape = factor(EC.char))) + 
  scale_shape_manual(values = c(1, 16)) +
  geom_jitter(width = 0.05, height = 0, size = 7, colour="blue") + 
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) + 
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Eurasian Coot") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

max_y <- max(comdataRTA.2$Black.winged.Stilt)
BS <- ggplot(comdataRTA.2, aes(x = Period, y = Black.winged.Stilt, alpha = 0.2, shape = factor(BS.char))) + 
  scale_shape_manual(values = c(1, 16)) +
  geom_jitter(width = 0.05, height = 0, size = 7, colour="blue") + 
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) +  
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Black-winged Stilt") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

max_y <- max(comdataRTA.2$Grey.Teal)
GT <- ggplot(comdataRTA.2, aes(x = Period, y = Grey.Teal, alpha = 0.2, shape = factor(GT.char))) + 
  scale_shape_manual(values = 16) +
  geom_jitter(width = 0.05, height = 0, size = 7, colour="blue") + 
  scale_x_discrete(limits = c("Reference","1996","2017","2021/22")) + 
  theme(legend.position = "none") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 14)) + 
  coord_cartesian(ylim = c(0, max_y*1.1)) +
  ylab("Abundance") + 
  ggtitle("Grey Teal") + 
  theme(plot.title = element_text(size = 16)) + 
  theme(axis.title.x = element_text(margin = margin(t = 10)), axis.title.y = element_text(margin = margin(r = 10)))

grid.arrange(FD, PD, AS, GT, EC, BS, ncol = 2, nrow = 3)
```

## Scatterplots and boxplots of guild richness by guild

```{r}
guilds <- read.csv("./Newdata/ToolibinGuilds.csv") #import Stuart Halse guild data
comdataRTA.guilds <- rawdata[rawdata$All_Ref %in% c("Y","M"), ] #copy raw data and extract only reference and modern surveys
comdataRTA.guilds[is.na(comdataRTA.guilds)]<-0 # replace na with 0
comdataRTA.guilds <- comdataRTA.guilds[grep("^T", row.names(comdataRTA.guilds)),] #extracts Toolibin
comdataRTA.guilds <- comdataRTA.guilds[, -(c(1,2,3,4,5,7,8))] #removes factors other than 'reference' and 'modern' surveys
comdataRTA.guilds$survey <- row.names(comdataRTA.guilds)
comdataRTA.guilds <- reshape2::melt(comdataRTA.guilds) #create long-format file from matrix
comdataRTA.guilds <- comdataRTA.guilds %>% rename_at('variable', ~'Species')
comdataRTA.guilds <- merge(x=comdataRTA.guilds, y=guilds, by="Species", all.x=TRUE);
comdataRTA.guilds <- as.data.table(comdataRTA.guilds[comdataRTA.guilds$value != 0, ]) #remove zero counts
write.csv(comdataRTA.guilds, "./guilds-species.csv")
#sum species richness by guild
comdataRTA.guilds <- comdataRTA.guilds %>% group_by(survey,All_Ref,Guild) %>% summarise(no_rows = length(Guild))
#change column headings
colnames(comdataRTA.guilds) <- c("Survey","Period","Guild","Richness")
#change values for Period
comdataRTA.guilds[comdataRTA.guilds == "Y"] <- "Reference"
comdataRTA.guilds[comdataRTA.guilds == "M"] <- "1996-2022"
#create short-format matrix to add in zero values for missing guilds in each survey
comdataRTA.guilds.m <- dcast(comdataRTA.guilds, formula = Survey ~ Guild, value="Richness", fill=0)
write.csv(comdataRTA.guilds.m, "./guildmatrix.csv")
#convert back to long-format
comdataRTA.guilds.2 <- as.data.table(reshape2::melt(comdataRTA.guilds.m)) #create long-format file from matrix
#add back the Period column by matching back to the original data file
colnames(comdataRTA.guilds.2)[2] <- 'Guild'
colnames(comdataRTA.guilds.2)[3] <- 'Richness'
#add Period back in to datafile
comdataRTA.guilds.2$Period <- comdataRTA.guilds$Period[match(comdataRTA.guilds.2$Survey, c(comdataRTA.guilds$Survey))]
#create new column for values of alpha (transparency) in ggplot
comdataRTA.guilds.2[, alphaprop := ifelse(comdataRTA.guilds.2$Period == "Reference", 0.4, 0.8)]
```

```{r  fig.align = 'center'}}
ggplot(comdataRTA.guilds.2, aes(x = Period, y = Richness, alpha = 0.4)) +
  geom_point(size=6, pch=16, colour="blue", alpha = ifelse(comdataRTA.guilds.2$Period == "Reference", 0.25, 0.5)) + scale_x_discrete(limits = c("Reference", "1996-2022")) + theme(legend.position = "none") +
  scale_alpha_continuous(range = c(0.2, 0.8)) + facet_wrap(vars(Guild))

ggplot(comdataRTA.guilds.2, aes(x = Period, y = Richness, alpha = alphaprop)) +
  geom_violin(colour="blue", fill="lightblue") + scale_x_discrete(limits = c("Reference", "1996-2022")) + theme(legend.position = "none") + facet_wrap(vars(Guild))

ggplot(comdataRTA.guilds.2, aes(Period, Richness)) +
  geom_boxplot(outlier.shape = NA, coef = 0) + scale_x_discrete(limits = c("Reference", "1996-2022")) +
  geom_jitter(width = 0.1, height = 0.1, size=2, pch=16, colour="blue", alpha = ifelse(comdataRTA.guilds.2$Period == "Reference", 0.5, 0.25)) + theme(legend.position = "none") +
  facet_wrap(vars(Guild))


ggplot(my_data, aes(x = x, y = y)) +
  geom_point(alpha = ifelse(my_data$x == 1, 0.5, 0.25)) +
  labs(title = "Custom Alpha for Different X-axis Values",
       x = "X-axis",
       y = "Y-axis")

```

## Heatmap of species abundances for Walbyring
```{r  fig.align = 'center'}}
comdataHMW <- rawdata[grep("^W", row.names(rawdata)),]
comdataHMW <- comdataHMW[comdataHMW$RICHNESS >4, ]
comdataHMW[is.na(comdataHMW)] <- 0 # replace na with 0
row.names(comdataHMW) <- comdataHMW$VISIT_DATE
comdataHMW <- comdataHMW[, -(1:8)]
comdataHMW <- comdataHMW[, colSums(comdataHMW) != 0]
comdataHMW.hel <-decostand(comdataHMW,"hellinger") # create helinger transformed matrix
melt_comdataHMW.hel <- melt(as.matrix(comdataHMW.hel))
ggplot(melt_comdataHMW.hel, aes(Var2, Var1)) + geom_tile(aes(fill = value), colour = "white") + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_gradient(low = "lightblue", high = "red") + labs(x="Species",y="Survey") + theme(axis.text.y = element_text(colour = c("black", "black","black","black","black","black","black","black","black","black","black","black","black","black","black","black","black", "blue", "blue", "blue", "blue", "blue", "blue")))
```


## Simper analysis of just the reference surveys versus 1996+ surveys for Lake Walbyring
```{r}
comdataRWA <- rawdata[rawdata$All_Ref %in% c("N","M"), ]
comdataRWA <- comdataRWA[grep("^W", row.names(comdataRWA)),] #extracts Walbyring
comdataRWA <- comdataRWA[comdataRWA$RICHNESS >4, ]
comdataRWA[is.na(comdataRWA)]<-0 # replace na with 0
comdataRWA.fact <- comdataRWA[, 1:8]
comdataRWA <- comdataRWA[, -(1:8)] #remove factors from data file
comdataRWA <- comdataRWA[, colSums(comdataRWA) != 0]
comdataRWA.hel<-decostand(comdataRWA,"hellinger") # create helinger transformed matrix
#simper on transformed data
comdataRWA.Simper <- simper(comdataRWA.hel, comdataRWA.fact$All_Ref)
summary(comdataRWA.Simper)
#simper on untransformed data
comdataRWA.Simper <- simper(comdataRWA, comdataRWA.fact$All_Ref)
summary(comdataRWA.Simper)
```


# INVERTEBRATE ANALYSES

## Multivariate analysis of composition
```{r include=FALSE}
rawdata.I.orig <- read.csv("./Newdata/Invert matrix for analyses2.csv", row.names = 1)
rawdata.I <- rawdata.I.orig[, 2:31] #use only Toolibin-Walbyring data (exclude SPS sites and species names)
rawdata.I <- t(rawdata.I)
row.names(rawdata.I) <- gsub("[.]", "-", row.names(rawdata.I))
Inv.fact.orig <- read.csv("./Newdata/Invert factors.csv", row.names = 1)
Inv.fact <- Inv.fact.orig[1:30, ]
#rawdata.I[is.na(rawdata.I)]<-0 # replace na with 0
```

```{r include=FALSE}
mds1.I <- metaMDS(rawdata.I, distance ="bray", k=2) #nMDS in 2 dimensions
#Extract mds2 coordinates and create new dataframe with coordinates, survey, and survey period
mds1.I.pts <- as.data.frame(mds1.I$points) #extract 2D ordination coordinates (points)
mds1.I.pts$Sample <- row.names(mds1.I.pts)
row.names(mds1.I.pts) <-gsub("[.]", "-", row.names(mds1.I.pts))
#mds1.I.pts$Year <- as.factor(Inv.fact[match(row.names(mds1.I.pts), row.names(Inv.fact)), 1])
mds1.I.pts$Period <- as.factor(Inv.fact[match(row.names(mds1.I.pts), row.names(Inv.fact[1:30, ])), 1])
mds1.I.pts$Survey <- as.factor(Inv.fact[match(row.names(mds1.I.pts), row.names(Inv.fact[1:30, ])), 5])
mds1.I.pts$Wetland <- as.factor(Inv.fact[match(row.names(mds1.I.pts), row.names(Inv.fact[1:30, ])), 3])
mds1.I.pts$Order <- as.factor(Inv.fact[match(row.names(mds1.I.pts), row.names(Inv.fact[1:30, ])), 4])
```

Below is a plot of a 2D ordination of invertebrate samples based on
species presence/absence data, with acceptable stress of
`r round(mds1.I$stress, digits=2)`.This shows the partially developed
community of April 2017 (A17) in both Toolibin and Walbyring (far right
of plot) being very different from other samples collected after longer
periods of inundation. This probably reflects an incompletely developed
community, since samples were collected only 2 months after the fill
event, but maybe also poor water quality and the time of year (see Cale
and Pinder, 2019). This sampling event is useful as an indicator of what
was present early following a summer fill event but does not represent
the contemporary invertebrate values of the lake. All of the Oct 2017,
Nov 2021 and Jan 2022 samples are located closer to, but still separate
from, the 1996 samples. Within the 2021-22 samples, the Nov samples
tended to group separate from the Jan samples suggesting some seasonal
turnover of species composition within the year. The most likely
explanation of the 1996 samples grouping separately to the 2017/21/22
spring/summer samples is the much high salinity (9400 mg/L) in Dec 1996
compared to the freshwater present in Oct 2017 (780), Nov 2021 (1770
mg/L) and Jan 2022 (2000 mg/L). It could be said that the 1996
community, being sampled when the lake was quite saline, was not typical
of invertebrate communities present in what is otherwise considered to
be a freshwater lake.

In three of the five sampling periods the Walbyring samples group
somewhat separate to the Toolibin samples, but for Apr 2017 and November
2021 this is not the case.

```{r fig.align = 'center'}
#first set axis limits based on maximum values of coordinates
Xmin <- min(mds1.I.pts$MDS1)*1.2 
Xmax <- max(mds1.I.pts$MDS1)*1.2
Ymin <- min(mds1.I.pts$MDS2)*1.2
Ymax <- max(mds1.I.pts$MDS2)*1.2
ggplot(mds1.I.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Wetland, pch = Survey), size=5) + coord_fixed(ratio = 1) + theme(text = element_text(size=15))
#+ geom_text(aes(label=Sample), size=2, vjust=1, hjust=1.3) #add labels if needed
```

## Multivariate analyses including five reference sites from Pinder, A.M., S.A. Halse, J.M. McRae, and R. J. Shiel. ‘Aquatic Invertebrate Assemblages of Wetlands and Rivers in the Wheatbelt Region of Western Australia’. Records of the Western Australian Museum Supplement 67 (2004): 7–37.


```{r include=FALSE}
#create datasets
rawdata.I2 <- rawdata.I.orig[, 2:36] #remove species names
rawdata.I2 <- t(rawdata.I2)
row.names(rawdata.I2) <- gsub("[.]", "-", row.names(rawdata.I2))
Inv.fact.orig <- read.csv("./Newdata/Invert factors.csv", row.names = 1) #import site factor variables
Inv.fact2 <- Inv.fact.orig
#rawdata.I[is.na(rawdata.I)]<-0 # replace na with 0
```

```{r include=FALSE}
#ordination
mds2.I <- metaMDS(rawdata.I2, distance ="bray", k=2) #nMDS in 2 dimensions
#Extract mds2 coordinates and create new dataframe with coordinates, survey, and survey period
mds2.I.pts <- as.data.frame(mds2.I$points) #extract 2D ordination coordinates (points)
mds2.I.pts$Sample <- row.names(mds2.I.pts)
row.names(mds2.I.pts) <-gsub("[.]", "-", row.names(mds2.I.pts))
#mds2.I.pts$Year <- as.factor(Inv.fact[match(row.names(mds2.I.pts), row.names(Inv.fact)), 1])
mds2.I.pts$Period <- as.factor(Inv.fact2[match(row.names(mds2.I.pts), row.names(Inv.fact2)), 1])
mds2.I.pts$Survey <- as.factor(Inv.fact2[match(row.names(mds2.I.pts), row.names(Inv.fact2)), 5])
mds2.I.pts$Wetland <- as.factor(Inv.fact2[match(row.names(mds2.I.pts), row.names(Inv.fact2)), 3])
mds2.I.pts$Order <- as.factor(Inv.fact2[match(row.names(mds2.I.pts), row.names(Inv.fact2)), 4])
```

```{r fig.align = 'center'}
#ordination plot
#first set axis limits based on maximum values of coordinates
Xmin <- min(mds2.I.pts$MDS1)*1.2 
Xmax <- max(mds2.I.pts$MDS1)*1.2
Ymin <- min(mds2.I.pts$MDS2)*1.2
Ymax <- max(mds2.I.pts$MDS2)*1.2
ggplot(mds2.I.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Wetland, pch = Survey), size=5) + coord_fixed(ratio = 1) + theme(text = element_text(size=15))
#+ geom_text(aes(label=Sample), size=2, vjust=1, hjust=1.3) #add labels if needed
```

```{r include=FALSE}
#re-run ordination without the April 2017 samples
rawdata.I3 <- rawdata.I.orig[, 2:36] #remove species names
rawdata.I3 <- rawdata.I3[, -grep("A17", colnames(rawdata.I3))] #exclude april 2017 columns
rawdata.I3 <- t(rawdata.I3)
row.names(rawdata.I3) <- gsub("[.]", "-", row.names(rawdata.I3))
Inv.fact.orig <- read.csv("./Newdata/Invert factors.csv", row.names = 1) #import site factor variables
Inv.fact3 <- Inv.fact.orig
#rawdata.I[is.na(rawdata.I)]<-0 # replace na with 0
```

```{r include=FALSE}
#ordination
mds3.I <- metaMDS(rawdata.I3, distance ="bray", k=2) #nMDS in 2 dimensions
#Extract mds2 coordinates and create new dataframe with coordinates, survey, and survey period
mds3.I.pts <- as.data.frame(mds3.I$points) #extract 2D ordination coordinates (points)
mds3.I.pts$Sample <- row.names(mds3.I.pts)
row.names(mds3.I.pts) <-gsub("[.]", "-", row.names(mds3.I.pts))
#mds3.I.pts$Year <- as.factor(Inv.fact[match(row.names(mds3.I.pts), row.names(Inv.fact)), 1])
mds3.I.pts$Period <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 1])
mds3.I.pts$Survey <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 5])
mds3.I.pts$Wetland <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 3])
mds3.I.pts$Order <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 4])
```


```{r include=FALSE}
#stress of the above ordination is very high (best 0.19) so repeated as a 3D ordination
mds3.I <- metaMDS(rawdata.I3, distance ="bray", k=3) #nMDS in 2 dimensions
#Extract mds2 coordinates and create new dataframe with coordinates, survey, and survey period
mds3.I.pts <- as.data.frame(mds3.I$points) #extract 2D ordination coordinates (points)
mds3.I.pts$Sample <- row.names(mds3.I.pts)
row.names(mds3.I.pts) <-gsub("[.]", "-", row.names(mds3.I.pts))
#mds3.I.pts$Year <- as.factor(Inv.fact[match(row.names(mds3.I.pts), row.names(Inv.fact)), 1])
mds3.I.pts$Period <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 1])
mds3.I.pts$Survey <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 5])
mds3.I.pts$Wetland <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 3])
mds3.I.pts$Order <- as.factor(Inv.fact2[match(row.names(mds3.I.pts), row.names(Inv.fact2)), 4])
mds3.I.pts$Col <- as.character(mds3.I.pts$Wetland)
mds3.I.pts$Col <- replace(mds3.I.pts$Col, mds3.I.pts$Col == "Toolibin", "red")
mds3.I.pts$Col <- replace(mds3.I.pts$Col, mds3.I.pts$Col == "Walbyring", "blue")
mds3.I.pts$Col <- replace(mds3.I.pts$Col, mds3.I.pts$Col == "Reference", "green")
mds3.I.pts$Col <- as.factor(mds3.I.pts$Col)
```

```{r fig.align = 'center'}
#ordination plot
#first set axis limits based on maximum values of coordinates
Xmin <- min(mds3.I.pts$MDS1)*1.2 
Xmax <- max(mds3.I.pts$MDS1)*1.2
Ymin <- min(mds3.I.pts$MDS2)*1.2
Ymax <- max(mds3.I.pts$MDS2)*1.2
ggplot(mds3.I.pts, aes(x=MDS1, y=MDS2)) + xlim(Xmin,Xmax) + ylim(Ymin,Ymax) + geom_point(aes(colour=Wetland, pch = Survey), size=5) + coord_fixed(ratio = 1) + theme(text = element_text(size=15))
#+ geom_text(aes(label=Sample), size=2, vjust=1, hjust=1.3) #add labels if needed
```

```{r}
#create new mds axis scores all >1 so drop lines all go down
mds3.I.pts$MDS1B <- mds3.I.pts$MDS1 + 1.0
mds3.I.pts$MDS2B <- mds3.I.pts$MDS2 + 1.0
mds3.I.pts$MDS3B <- mds3.I.pts$MDS3 + 1.0

#convert Wetland factor to character so can replace values with numbers, then convert to numeric so the drop line calculations work with plot_ly. See https://community.plotly.com/t/droplines-from-points-in-3d-scatterplot/4113
mds3.I.pts$wetland <- as.character(mds3.I.pts$Wetland)
mds3.I.pts['wetland'][mds3.I.pts['wetland'] == 'Toolibin'] <- '1'
mds3.I.pts['wetland'][mds3.I.pts['wetland'] == 'Walbyring'] <- '2'
mds3.I.pts['wetland'][mds3.I.pts['wetland'] == 'Reference'] <- '3'
mds3.I.pts$wetlands <- as.numeric(mds3.I.pts$wetland)

#Duplicate dataset and create new mds3 with 0 values so drop lines go to zero
mds3.I.pts$id <- seq_len(nrow(mds3.I.pts))
ms <- replicate(2, mds3.I.pts, simplify = F)
ms[[2]]$MDS3B <- 0
m <- ms %>%
  bind_rows() %>%
  group2NA("id", "wetland")

plot_ly(color = ~factor(wetland), colors = c("orange", "blue", "red"), showlegend = T) %>%
  add_markers(data = mds3.I.pts, x = ~MDS1B, y = ~MDS2B, z = ~MDS3B) %>%
  add_paths(data = m, x = ~MDS1B, y = ~MDS2B, z = ~MDS3B) %>%
  layout(
    xaxis = list(
      showline = TRUE, # Show x-axis line
      linecolor = 'black', # Color of x-axis line
      linewidth = 2, # Width of x-axis line
      gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
      gridwidth = 1 # Width of gridlines
    ),
    yaxis = list(
      showline = TRUE, # Show y-axis line
      linecolor = 'black', # Color of y-axis line
      linewidth = 2, # Width of y-axis line
      gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
      gridwidth = 1 # Width of gridlines
    ),
    zaxis = list(
      showline = TRUE, # Show z-axis line
      linecolor = 'black', # Color of z-axis line
      linewidth = 2, # Width of z-axis line
      gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
      gridwidth = 1 # Width of gridlines
    ),
    scene = list(
      xaxis = list(
        showline = TRUE, # Show x-axis line
        linecolor = 'black', # Color of x-axis line
        linewidth = 2, # Width of x-axis line
        gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
        gridwidth = 1 # Width of gridlines
      ),
      yaxis = list(
        showline = TRUE, # Show y-axis line
        linecolor = 'black', # Color of y-axis line
        linewidth = 2, # Width of y-axis line
        gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
        gridwidth = 1 # Width of gridlines
      ),
      zaxis = list(
        showline = TRUE, # Show z-axis line
        linecolor = 'black', # Color of z-axis line
        linewidth = 2, # Width of z-axis line
        gridcolor = 'rgba(0, 0, 0, 0.5)', # Darker gridlines color
        gridwidth = 1 # Width of gridlines
      )
    )
  )
```

## cluster analyses by composition for 1996 to 2022 samples (excluding April 2017)
```{r  fig.align = 'center'}}
clusdist <- vegdist(rawdata.I3, method="bray")
clusanal <- hclust(clusdist, method="average") 
plot(clusanal, xlab="Survey", ylab="Bray")
```


## Stacked invertebrate richness plots with samples seperate

```{r}
#Load and format data
#create long file for aggregation in next chunks
rawdata.IR.orig <- t(read.csv("./Newdata/Invert matrix for richness4.csv", row.names = 1))
rawdata.IR <- as.matrix(rawdata.IR.orig)
Inv.stack <- as.data.frame(as.table(rawdata.IR))
colnames(Inv.stack) <- c("Sample", "Code", "Presence")
Inv.stack$Wetland <- substr(Inv.stack$Sample, 1, 1) #create new variable for wetland (Toolibin or Walbyring)
Inv.stack$Period <- substr(Inv.stack$Sample, 4,6) #create new variable for survey period
Inv.stack$Period <- gsub("^(?!N|O|J|D|A).*", "ref", Inv.stack$Period, perl = TRUE) #substitute ref in the Period column
Inv.stack$Sample <-gsub("[.]", "-", Inv.stack$Sample) #convert dot in sample names to a dash for consistency between files (the dots are created on sample import).
Taxonomy3 <- read.csv("./Newdata/Taxonomy3.csv")
colnames(Taxonomy3) <- c("Group", "Group.order", "Code", "LowestID")
Inv.stack$Group <- Taxonomy3[match(Inv.stack$Code, Taxonomy3$Code), 1] #add invertebrate group to the data field to be used for stacking.
Inv.stack$Presence <- as.numeric(Inv.stack$Presence)
Inv.fact.orig <- read.csv("./Newdata/Invert factors.csv", row.names = 1)
```


```{r fig.align = 'center'}
#Tolibin plots by sample
# Aggregate data by taxonomic group and sample for just Lake Toolibin and produce a stacked column plot
Inv.stack.T <- subset(Inv.stack, Wetland %in% c("T","S")) #create Toolibin only dataset
Inv.stack.T.agg <- aggregate(data=Inv.stack.T, Presence ~ Sample + Group, FUN = "sum") #aggregate count data by species and survey
#Inv.stack.T.agg <- Inv.stack.T.agg[!Inv.stack.T.agg$Group %in% c("Nematodes","Turbellarians","Hydrozoans","Sponges"), ]
#Inv.fact.orig$Period <- substr(row.names(Inv.fact.orig), 4,6) #not needed now?
Inv.stack.T.agg$Order <- Inv.fact.orig[match(Inv.stack.T.agg$Sample, row.names(Inv.fact.orig)), 4] #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.T.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.T.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.T.agg <- setorder(Inv.stack.T.agg, cols = "Grouporder") #order by taxonomic group
Inv.stack.T.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.T.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
```

```{r  fig.align = 'center'}}
ggplot(Inv.stack.T.agg, aes(x=reorder(Sample, Order), y=Presence, fill=Grouporder)) + 
  geom_bar(stat="identity", colour = "black") +
  scale_fill_manual(values = c("#FFD700","#FFA54F","#CDB38B","#FFAEB9","#FFF68F","lavender","white","plum2","#BCEE68","#66CD00","darkcyan","#FFFF00","#7FFFD4","#9AC0CD","#79CDCD","#00FFFF","#98F5FF","deepskyblue2"), labels = c('sponges','hydrozoans','Turbellarians','Nematodes','snails','leeches','oligochaetes','water mites','cladocerans','ostracods','copepods','amphipods','beetles','dipterans','hemipterans','lepiodpterans','dragonflies+damselflies','caddisflies')) + 
  xlab("Sample") +
  ylab("Number of invertebrate species") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
  theme(axis.title.y = element_text(vjust = +2.5)) +
  theme(legend.title=element_blank()) + 
    geom_vline(xintercept = c(3.5, 6.5, 9.5, 12.5, 15.5), color = "black", linetype = "solid")
```

## Walbyring plots by sample
```{r fig.align = 'center'}
# Aggregate data by taxonomic group and sample for just Lake Walbyring and produce a stacked column plot
Inv.stack.W <- subset(Inv.stack, Wetland %in% c("W","S")) #create Toolibin only dataset
Inv.stack.W.agg <- aggregate(data=Inv.stack.W, Presence ~ Sample + Group, FUN = "sum") #aggregate count data by species and survey
#Inv.stack.W.agg <- Inv.stack.W.agg[!Inv.stack.W.agg$Group %in% c("Nematodes","Turbellarians","Hydrozoans","Sponges"), ]
#Inv.fact.orig$Period <- substr(row.names(Inv.fact.orig), 4,6) #not needed now?
Inv.stack.W.agg$Order <- Inv.fact.orig[match(Inv.stack.W.agg$Sample, row.names(Inv.fact.orig)), 4] #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.W.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.W.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.W.agg <- setorder(Inv.stack.W.agg, cols = "Grouporder") #order by taxonomic group
Inv.stack.W.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.W.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
```

```{r  fig.align = 'center'}}
ggplot(Inv.stack.W.agg, aes(x=reorder(Sample, Order), y=Presence, fill=Grouporder)) + 
  geom_bar(stat="identity", colour = "black") +
  scale_fill_manual(values = c("#FFD700","#FFA54F","#CDB38B","#FFAEB9","#FFF68F","lavender","white","plum2","#BCEE68","#66CD00","darkcyan","#FFFF00","#7FFFD4","#9AC0CD","#79CDCD","#00FFFF","#98F5FF","deepskyblue2"), labels = c('sponges','hydrozoans','Turbellarians','Nematodes','snails','leeches','oligochaetes','water mites','cladocerans','ostracods','copepods','amphipods','beetles','dipterans','hemipterans','lepiodpterans','dragonflies+damselflies','caddisflies')) +
  xlab("Sample") +
  ylab("Number of invertebrate species") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
  theme(axis.title.y = element_text(vjust = +2.5)) +
  theme(legend.title=element_blank()) +
  geom_vline(xintercept = c(3.5, 6.5, 9.5, 12.5, 15.5), color = "black", linetype = "solid")
```

## Stacked invertebrate richness plots with samples grouped for each sampling period

```{r}
#Load and format data
#create long file for aggregation in next chunks
rawdata.IR.orig <- t(read.csv("./Newdata/Invert matrix for richness by period2.csv", row.names = 1)) #load seperate datafile with occurrence lumped to sampling period and adjusted for larvae etc.
rawdata.IR <- as.matrix(rawdata.IR.orig)
Inv.stack <- as.data.frame(as.table(rawdata.IR))
colnames(Inv.stack) <- c("Period", "Code", "Presence")
Inv.stack$Wetland <- substr(Inv.stack$Period, 1, 1) #create new variable for wetland (Toolibin or Walbyring)
Inv.stack$Period <-gsub("[.]", "-", Inv.stack$Period) #convert dot in sample names to a dash for consistency between files (the dots are created on sample import).
Inv.stack$Period <- substr(Inv.stack$Period, 3,5) #create new variable for survey period
#Inv.stack$Period <- gsub("^(?!N|O|J|D|A).*", "ref", Inv.stack$Period, perl = TRUE) #substitute ref in the Period column
Taxonomy3 <- read.csv("./Newdata/Taxonomy3.csv")
colnames(Taxonomy3) <- c("Group", "Group.order", "Code", "LowestID")
Inv.stack$Group <- Taxonomy3[match(Inv.stack$Code, Taxonomy3$Code), 1] #add invertebrate group to the data field to be used for stacking.
Inv.stack$Presence <- as.numeric(Inv.stack$Presence)
Inv.fact.orig <- read.csv("./Newdata/Invert factors.csv", row.names = 1)
```

```{r fig.align = 'center'}
#plot for Toolibin by group
# Aggregate data by taxonomic group and sample for just Lake Toolibin and produce a stacked column plot
Inv.stack.T.G <- subset(Inv.stack, Wetland %in% c("T")) #create Toolibin only dataset
Inv.stack.T.G.agg <- aggregate(data=Inv.stack.T.G, Presence ~ Period + Group, FUN = "sum") #aggregate count data by species and survey
#Inv.stack.T.G.agg <- Inv.stack.T.G.agg[!Inv.stack.T.G.agg$Group %in% c("Nematodes","Turbellarians","Hydrozoans","Sponges"), ]
#Inv.fact.orig$Period <- substr(row.names(Inv.fact.orig), 4,6) #not needed now?
Inv.stack.T.G.agg$Order <- Inv.fact.orig[match(Inv.stack.T.G.agg$Period, Inv.fact.orig$Survey), 4] #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.T.G.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.T.G.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.T.G.agg <- setorder(Inv.stack.T.G.agg, cols = "Grouporder") #order by taxonomic group
Inv.stack.T.G.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.T.G.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
```

```{r  fig.align = 'center'}}
ggplot(Inv.stack.T.G.agg, aes(x=reorder(Period, Order), y=Presence, fill=Grouporder)) + 
  geom_bar(stat="identity", colour = "black") +
  scale_fill_manual(values = c("#FFA54F","#CDB38B","#FFAEB9","#FFF68F","lavender","white","plum2","#BCEE68","#66CD00","darkcyan","#FFFF00","#7FFFD4","#9AC0CD","#79CDCD","#00FFFF","#98F5FF","deepskyblue2"), labels = c('hydrozoans','Turbellarians','Nematodes','snails','leeches','oligochaetes','water mites','cladocerans','ostracods','copepods','amphipods','beetles','dipterans','hemipterans','lepiodpterans','dragonflies+damselflies','caddisflies')) + 
  xlab("Sample") +
  ylab("Number of invertebrate species") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
  theme(axis.title.y = element_text(vjust = +2.5)) +
  theme(legend.title=element_blank())
```

```{r}
#plot for Walbyring by group
# Aggregate data by taxonomic group and sample for just Lake Toolibin and produce a stacked column plot
Inv.stack.W.G <- subset(Inv.stack, Wetland %in% c("W")) #create Toolibin only dataset
Inv.stack.W.G.agg <- aggregate(data=Inv.stack.W.G, Presence ~ Period + Group, FUN = "sum") #aggregate count data by species and survey
#Inv.stack.W.G.agg <- Inv.stack.W.G.agg[!Inv.stack.W.G.agg$Group %in% c("Nematodes","Turbellarians","Hydrozoans","Sponges"), ]
#Inv.fact.orig$Period <- substr(row.names(Inv.fact.orig), 4,6) #not needed now?
Inv.stack.W.G.agg$Order <- Inv.fact.orig[match(Inv.stack.W.G.agg$Period, Inv.fact.orig$Survey), 4] #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.W.G.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.W.G.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
Inv.stack.W.G.agg <- setorder(Inv.stack.W.G.agg, cols = "Grouporder") #order by taxonomic group
Inv.stack.W.G.agg$Grouporder <- as.factor(Taxonomy3[match(Inv.stack.W.G.agg$Group, Taxonomy3$Group), 2]) #add in a variable for ordering the samples on the column plots- from the factor file.
```

```{r  fig.align = 'center'}}
ggplot(Inv.stack.W.G.agg, aes(x=reorder(Period, Order), y=Presence, fill=Grouporder)) + 
  geom_bar(stat="identity", colour = "black") +
  scale_fill_manual(values = c("#FFA54F","#CDB38B","#FFAEB9","#FFF68F","lavender","white","plum2","#BCEE68","#66CD00","darkcyan","#FFFF00","#7FFFD4","#9AC0CD","#79CDCD","#00FFFF","#98F5FF","deepskyblue2"), labels = c('hydrozoans','Turbellarians','Nematodes','snails','leeches','oligochaetes','water mites','cladocerans','ostracods','copepods','amphipods','beetles','dipterans','hemipterans','lepiodpterans','dragonflies+damselflies','caddisflies')) +  
  xlab("Sample") +
  ylab("Number of invertebrate species") + 
  theme(axis.text = element_text(size = 12)) + 
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
  theme(axis.title.y = element_text(vjust = +2.5)) +
  theme(legend.title=element_blank())
```
